---
title: "Homework 1: Graphical Exploration with GGplot and Numerical Summaries - Biostatistics I"
author: "Ryan Duggan"
date: "September 15, 2018"
output: pdf_document
---

```{r echo=FALSE}
knitr::opts_chunk$set(cache=TRUE, error=TRUE, message=TRUE, warning=TRUE)
#loading tidyverse + other necessary packages 
library(tidyverse)
library(foreign)
library(knitr)
library(kableExtra)
library(gridExtra)

#reading nhgh dataset into global environment 
nhgh <- read.dta(file = "../data/nhgh.dta")
#Recoding variables - these two functions accomplish the same thing...
#nhgh$tx <- as.factor(plyr::mapvalues(nhgh$tx, from = c(0,1), to = c("No Treatment", "On Insulin or Diabetes Meds")))

nhgh$tx <- factor(nhgh$tx, levels = c(0:1), labels = c("No Treatment", "On Insulin or Diabetes Meds"))
nhgh$dx <- factor(nhgh$dx, levels = c(0:1), labels = c("No Diagnosis", "Diagnosed with DM or Pre-DM"))
```
# Question 1 - BMI

## a) Make a histogram of the BMI variable. b) Make a density plot of the BMI variable.
```{r echo = T, fig.width = 10}

plot1 <- ggplot(data = nhgh)+
  geom_histogram(aes(x = bmi), binwidth = 1)+
  ggtitle("Histogram of Patient BMI")+
  labs(x = "BMI", y = "Count")+
  theme(plot.title = element_text(hjust = 0.5))

plot2 <- ggplot(data = nhgh)+
  geom_density(aes(x = bmi), fill = "gray")+
  ggtitle("Density Plot of Patient BMI")+
  labs(x = "BMI", y = "Density")+
  theme(plot.title = element_text(hjust = 0.5))

grid.arrange(plot1, plot2, nrow = 1)

```

## c) Write one sentence summarizing the trends seen in the figures.
Patient BMI in this dataset shows a normal distribution around a mean of 28.32 (sd = 6.95), with a slight right skew. This trend is evident in both the histogram and density plot above.

## d) Is the histogram or the density plot better in this case? Explain
There do not appear to be any major non-aesthetic differences between the density plot and the histogram of BMI. This makes sense, as density plots measure distribution in the same way as the histogram, but with the application of a Gaussian filter to smooth the plot.

&nbsp;&nbsp;&nbsp;&nbsp;

# Question 2 - Glycohemoglobin

## a) Make a histogram of the glycohemoglobin data. b) Make a density plot of the glycohemoglobin data.
```{r echo = T, fig.width = 10}
plot3 <- ggplot(data = nhgh)+
  geom_histogram(aes(x = gh), binwidth = 0.5)+
  ggtitle("Histogram of Patient Glycohemoglobin Percentage")+
  labs(x = "Glycohemoglobin Percentage", y = "Count")+
  theme(plot.title = element_text(hjust = 0.5))

plot4 <- ggplot(data = nhgh)+
  geom_density(aes( x = gh), fill = "gray")+
  ggtitle("Density Plot of Patient Glycohemoglobin Percentage")+
  labs(x = "Glycohemoglobin Percentage", y = "Density")+
  theme(plot.title = element_text(hjust = 0.5))

grid.arrange(plot3, plot4, nrow = 1)
```

## c) Make a density plot of the gylcohemoglobin data with a different curve for patients using insulin (tx = 1) and patients not using insulin (tx = 0).
For this question, you should use different colors for the different densities.
```{r echo = T, fig.width = 7.5, fig.align = "center"}
ggplot(data = nhgh, aes(x = gh, fill = tx))+
  geom_density(position = "stack", alpha = 0.5)+
  ggtitle("Density Plot of Gycohemoglobin (Gh) Percentage by Diabetes Treatment Status")+
  labs(x = "Gh Percentage", y = "Density")+
  theme(plot.title = element_text(hjust = 0.5))
```

## d) Describe the trends you see in the figures

These plots are showing the frequency of occurences of glycohemoglobin percentage in the patient sample broken down by whether they are receiving insulin treatment (tx = 1) or no treatment (tx = 0). Patients receiving treatment showed a higher mean and median Gh level than patients not receiving treatment. 

This pattern holds up when breaking down Gh by DM diagnosis instead of treatment type, which would be expected as Gh levels are correlated with the presence of diabetes, and presumably patients without diabetes would not be receiving treatment. (See below.)

```{r echo = T, fig.width = 7.5, fig.align = "center"}
group_by(nhgh, tx) %>% 
  summarize(., mean = mean(gh, na.rm = T), median = median(gh, na.rm = T), sd = sd(gh, na.rm = T)) %>%
  kable(col.names = c("Treatment Status", "Mean", "Median", "Standard Deviation")) %>%
  kable_styling(bootstrap_options = "striped")

ggplot(data = nhgh, aes(x = gh, fill = dx))+
  geom_density(position = "stack", alpha = 0.5)+
  ggtitle("Density Plot of Gycohemoglobin (Gh) Percentage by Diabetes Diagnosis")+
  labs(x = "Gh Percentage", y = "Density")+
  theme(plot.title = element_text(hjust = 0.5))
```

&nbsp;&nbsp;&nbsp;&nbsp;

# Question 3 - Race/Ethnicity

## a) Create a figure showing the relative numbers in each race/ethnicity within the insulin and no insulin groups
```{r echo = T, fig.width = 7.5, fig.height = 5, fig.align = "center"}
ggplot(data=nhgh, aes(x = tx, fill = re))+
  geom_bar(position = "fill")+  #this makes the bars proportional instead of absolute
  ggtitle("Bar Graph of Diabetes Treatment Status by Race")+
  labs(x = "Diabetes Treatment", "Proportion", fill = "Race")
```

## b) Summarize what you see in the figure.
This graph shows that roughly equal (or at least very similar) proportions of each racial group receive each treatment type, with minority (non-white) groups tending to have a higher proportion receiving treatment than not. 

&nbsp;&nbsp;&nbsp;&nbsp;

# Question 4 -  BMI and race

## a) Make side-by-side boxplots of BMI for each race.
```{r echo=T, fig.width= 7.5, fig.align = "center"}
ggplot(data = nhgh, aes(x=re, y = bmi))+
  geom_boxplot()+
  ggtitle("Boxplot of BMI by Race")+
  labs(x = "", y = "BMI")+
  theme(plot.title = element_text(hjust = 0.5))
```

## b) Create a table that contains each of the following values of the BMI for each race: mean, median, min, max, standard deviation, and IQR
```{r echo = T, fig.width = 7.5, fig.align = "center"}
nhghByRace <- group_by(nhgh, re)
nhghByRace %>%
  summarize(., mean = mean(bmi), median = median(bmi), sd = sd(bmi), IQR = IQR(bmi), min = min(bmi), max = max(bmi))%>%
  kable(col.names = c("Race", "Mean", "Median", "Standard Deviation", "IQR", "Minimum", "Maximum")) %>%
  kable_styling(bootstrap_options = "striped")
```

## c) Describe how BMI differs among the races
Although test for statistical significance were not carried out, the boxplots and summary statistics indicate that average BMI is similar across different races, with the highest mean and median being found in the Non-Hispanic Black group, and the lowest being the "Other Race" group. Aside from this difference, BMI seems to be normally distributed in all racial groups.

&nbsp;&nbsp;&nbsp;&nbsp;

# Question 5 - BMI, glycohemoglobin, and insulin

## a) Make a scatter plot of gylcohemoglobin versus BMI colored by insulin use

```{r echo = T, fig.width = 7.5, fig.align = "center"}
ggplot(data = nhgh)+
  geom_point(aes(x = bmi, y = gh, color = tx), size = 2, alpha = 0.5)+
  ggtitle("Scatterplot of BMI and Gh Measurement by Diabetes Treatment Status")+
  scale_color_discrete(name = "Insulin Treatment")
```

## b) Describe what you see in the plot.
There appears to be a very strong clustering of lower BMI and lower Gh percentages in patients not being treated for diabetes, with even otherwise healthy patients with higher BMIs tending to have low Gh levels. Patients who are receiving insulin or other treatments have a wider spread of Gh levels, with a tendency to cluster at a slightly higher level compared to the No Treatment group, while also tending to remain in a similar range of BMI.

