---
title: "Biostatistics I - Assignment 2"
author: "Ryan C Duggan"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: pdf_document
---
```{r echo=T}
knitr::opts_chunk$set(cache=TRUE, error=TRUE, message=TRUE, warning=TRUE)
#loading tidyverse + other necessary packages 
library(tidyverse)
```

## Question 1: Blood types probability
# The following table shows the prevalence of ABO block groups for the U.S. Caucasian and American Black populations. The table specifies the probabilities of a specific race and blood type. For example, the probability that a randomly selected individual is black with blood type O is 0.098.


|                |    A     |    B     |     AB  |      O    |
| -------------- |:--------:|:--------:|:--------:|:--------:|
| Caucasian      |   0.352  |   0.064  |   0.024  |  0.360   |
| American Black |   0.054  |   0.040  |   0.008  |  0.098   |

## Question 1.1
# What is the probability that a randomly selected individual does not have blood type AB?


The probability that a randomly-selected individual does not have blood type AB is 0.968, which is calculated by subtracting the probability of an individual having blood type AB from 1. (1- (0.024 + 0.008) = 0.968)



## Question 1.2
# What is the probability that two people selected at random both have blood type B?


The probability that two individuals selected at random both have blood type B is 0.0196.



## Question 1.3
# Are the events "blood type B" and "American black" race statistically independent?


These two events are not statistically independent, as an individual's ethnic group influences his or her blood type, and because of this genetic influence, it is impossible to separate the two. Furthermore, from a statistical perspective, if one knows that an individual is black, according to the table he or she has a probability of 0.04 of having type B blood. Since in this case race conveys information about blood type, the two events cannot be said to be statistically independent.



## Question 1.4
# Are the events "blood type O" and "blood type A" mutually exclusive?


These two events are mutually exclusive, as an individual cannot have two blood types at the same time. The chance of that occurring in a single individual is thus zero.



## Question 1.5
# Given an example of two events that are not mutually exclusive.


Two events that are not mutually exclusive are an individual being blood type A and being Caucasian.



## Question 1.6
# What is the conditional probability of "blood type A" given that the race is "American black"?


The conditional probability of an American black individual having blood type A is 0.27, which can be calculated according to the formula for conditional probability:
```{r}
0.054/(0.054+0.040+.008+.098)
```


## Question 2: Death probability
# The following data records the reported proportions of 20- to 25-year-old White males causes of death within a five year period. In other words, it gives the probability of death within five years.


|  Cause           |    Probability  |
| ---------------- |-----------------:|
| Suicide          |   0.00126        |   
| Homicide         |   0.00063        | 
| Auto accident    |   0.00581        |
| Leukemia         |   0.00023        |
| all other causes |   0.00788        |

## Question 2.1
# What is the probability that a White male aged 20 to 25 dying from any cause of death within the next five years?


All of these events are mutually exclusive, thus the probability of a white male aged 20-25 dying of any of the listed causes is the sum of their individual probabilities: 0.01581. 
```{r}
sum(0.00126 + .00063 + .00581 + .00023 + .00788)
```


## Question 2.2
# Out of 100,000 white males in the 20 to 25 age group, how many deaths would you expect in the next 5 years from leukemia?

One would expect there to be 0.00023 * 100,000 = 23 deaths from leukemia in the next five years.



## Question 2.3
# Given that a White male aged 20 to 25 years has died, what is the most likely cause of death? Assume that nothing else is known. Provide rationale for your answer.


Based on the probability table, the most likely cause would fall under "all other causes," since it has the highest probability. However, in terms of specific causes of death, the most likely would be auto accident, as it has the highest probability and is thus the most likely to occur.



## Question 3: Lab test one
# Suppose that a lab test values are normally distributed with $\mu = 80$ and $\sigma = 3$. 

## Question 3.1
# What is the probability that a randomly selected test has a value that is within one standard deviation of the mean? Round to three decimal places.


There are two ways of solving this - the first uses the 68-95-99 rule, which states that in a normal distribution, 68% of values fall within one standard deviation of the mean. This indicates that the probability a randomly-selected value falls within one standard deviation of the mean is ~0.68.

Alternatively, this could be calculated by finding the probability of the value lying between the mean and standard deviation and the mean - standard deviation P(mean - SD <= X <= mean + SD). Calculating this in R produces a probability of 0.683:

```{r}
(pnorm(83, mean = 80, sd = 3) - pnorm(77, mean = 80, sd = 3)) %>% 
  round(3)
```


## Question 3.2
# What is the probability that a randomly selected test has a value that is within two standard deviations of the mean? Round to three decimal places.

Similarly, using the 68-95-99 rule, it can be estimated that the probability of a randomly-selected value falling within two standard deviations of the mean would be around 0.95. By using the pnorm function to calculate P(mean - 2*SD <= X <= mean + 2*SD) confirms this by returning a probability of 0.954:

```{r}
(pnorm(86, mean = 80, sd = 3) - pnorm(74, mean = 80, sd = 3)) %>% 
  round(3)
```


## Question 3.3
# What is the probability that a randomly selected test has a value that is within three standard deviations of the mean? Round to three decimal places.


The probability in this case is P(mean - 3*SD <= X <= mean + 3*SD). Using the pnorm function produces the expected result of 0.997:

```{r}
(pnorm(89, mean = 80, sd = 3) - pnorm(71, mean = 80, sd = 3)) %>%
  round(3)
```


## Question 4: Lab test two
# Suppose that another lab test values record changes from a reference value and these are normally distributed with $\mu = -16$ and $\sigma = 1.5$. 

## Question 4.1
# What is the probability that a randomly selected test has a value that is within one standard deviation of the mean? Round to three decimal places.


The probability P(mean - SD <= X <= mean + SD) would be calculated in the same way as the previous question, producing a probability of 0.683:

```{r}
(pnorm(-14.5, mean = -16, sd = 1.5) - pnorm(-17.5, mean = -16, sd = 1.5)) %>% 
  round(3)
```


## Question 4.2
# What is the probability that a randomly selected test has a value that is within two standard deviations of the mean? Round to three decimal places.


The probability P(mean - 2*SD <= X <= mean + 2*SD) would be calculated in the same way as the previous question, producing a probability of 0.954:

```{r}
(pnorm(-13, mean = -16, sd = 1.5) - pnorm(-19, mean = -16, sd = 1.5)) %>% 
  round(3)
```


## Question 4.3
# What is the probability that a randomly selected test has a value that is within three standard deviations of the mean? Round to three decimal places.

The probability P(mean - 3*SD <= X <= mean + 3*SD) would result in a probability of 0.997:

```{r, echo = TRUE}
(pnorm(-11.5, mean = -16, sd = 1.5) - pnorm(-20.5, mean = -16, sd = 1.5)) %>% 
  round(3)
```


## Question 4.4
# How do your answers compare to those of Question 3? Does this surprise you?


The answers are identical - this is not surprising, as both samples are normally distributed, and thus all values tend to fall within each standard deviation in a pattern according to the 68-95-99 rule, meaning that regardless of the mean and standard deviation, the same proportion of the distribution would fall within x standard deviations of the mean.



## Question 5: Randomization verification
# A clinical trial was designed to determine whether radiation therapy or surgery was more effective in reducing disease recurrence in men with prostate cancer. Patients were randomized to each treatment in a 1:1 manner, which implies the chance a patient is randomized to either arm is 0.50. Suppose that in the first 20 patients randomized at a particular site participating in the trial, 17 patients were assigned to the surgery arm. The site complains that the randomization does not appear to be working. Does the site's suspicion have merit? 

```{block,  box.title = "Tip", box.body = list(fill = "lemonchiffon")}
When you answer this, you should consider all scenarios that would support the suspicions that the randomization is broken. Specifically, what other outcomes out of 20 would be more extreme to the one observed. You might want to calculate the probability of getting the result observed or one more extreme.
```


Since randomization has a binary outcome, this can be calculated with the binomial distribution function pbinom to find the probability P(X > 16), resulting in a probability of 0.00129. This indicates that it is extremely unlikely that the patients were randomized correctly.

```{r}
pbinom(16, 20, 0.5, lower.tail = FALSE) %>%
  round(5)
```


## Hypertension
# The remainder of the questions are relevant to this scenario. 

# The proportion of U.S. adults over the age of 20 with hypertension is 0.335, or 33.5%. 

## Question 6: Hypertension
# Suppose you take a random sample of 64 U.S. adults over the age of 20.

## Question 6.1
# What is the chance that 32 people have hypertension? (Round to 4 decimal places.)


The probability that exactly 32 out of 64 people in this sample have hypertension is 0.0025.

```{r}
dbinom(32, 64, 0.335) %>%
  round(4)
```

## Question 6.2
# What is the chance that more than 20 individuals in the sample have hypertension? (Round to 3 decimal places.)


The probability that more than 20 individuals in the sample have hypertension is 0.593.

```{r}
pbinom(q = 20, size = 64, prob = 0.335, lower.tail = F) %>%
  round(3)
```



## Question 6.3
# What is the chance that between 21 and 24 individuals, inclusive, out of 64 have hypertension? (Round to 3 decimal places.)


The probability P(21 <= X <= 24) is equal to 0.385.

```{r}
(pbinom(24, 64, 0.335) - pbinom(20, 64, 0.335)) %>%
  round(3)
```


## Question 7: Hypertension
# Now suppose you take your random sample (of size 64) from one city in the United States. You get a sample proportion of 0.211.


## Question 7.1
# How likely are you to get a sample proportion this small or smaller if this city is representative of the US population?


#Given the sample proportion p = 0.211, solving for y = p * n results in a y (number of observations / successes) of 13.5. Modeling this in R produces a probability of 0.015.

```{r}
pbinom(q = 13.5, size = 64, prob = 0.335) %>%
  round(3)
```


## Question 7.2
# What do you conclude?


Given that the probability of obtaining this sample proportion is so low, it is very unlikely that such a sample proportion would be obtained if the city is representative of the US population.



## Question 8: Hypertension
# Give a range for which you are 95% certain would contain the sample proportion of adults with hypertension if you were to take a random sample of 64 adults in the U.S. over the age of 20. (Round to three decimal places.)


The range that we are 95% sure contains the sample proportion is 0.219 - 0.453.

```{r}
#Lower limit of CI 
(qbinom(0.025, 64, 0.335)/64) %>%
  round(3)
#Upper limit of CI (1 - 0.025) 
(qbinom(0.975, 64, 0.335)/64) %>%
  round(3)
```


## Question 9: Hypertension
# We want to compare the sampling distribution of our sample proportion to that of a normal distribution. The normal distribution should have the same mean and standard deviation as the sampling distribution. To do this, we want to use `qqplot()`. We will simulate the sampling distribution of both the sampling proportion as well as the normal distribution. After using `qqplot()`, use `qqline()` to add a line to the plot.

## Question 9.1
# Make a qqplot comparing the sampling distribution of the sample proportion to that of a normal distribution, including a line that indicates a perfect relationship. 

```{r}
# creating normal distribution for comparison
HTnorm <- rnorm(n = 100, mean = 0.335, sd = 0.0589)

# creating binomial distribution using hypertension question parameters

HTbinom <- (rbinom(n = 100, size = 64, prob = 0.335)/64)

#generating QQplot + qqline
qqplot(x = HTnorm, y = HTbinom, main = "Q-Q Plot of Hypertension Data Quantiles", 
       xlab = "Normal Quantiles", ylab = "Hypertension Sampling Quantiles")
qqline(y = HTbinom, distribution = function(p) qnorm(p, 0.335, 0.0589))

```


## Question 9.2
# How do these two distributions compare?


The two distributions appear to be very similar, as evidenced by the tendency to cluster around the reference line. Since the comparison sample was taken from a normal distribution, the similarity of the hypertension sampling distribution indicates that the population being sampled is also normally-distributed.



## Question 9.3
# To verify your answer to Question 9.2, use a normal distribution with the same mean and standard deviation of the binomial distribution to determine the likelihood that we observe between 21 and 24 individuals, inclusive, out of 64 have hypertension (Question 6.3). How do these values compare?

```{r}
# Normal distribution with same mean and standard deviation as previous binomial 
# distribution for P(21 <= X <= 24):

#Calculating mean and SD of the binomial distribution ([1] = mean, [2] = SD):
HTnormVals <- c((64 * 0.335), (sqrt(64*0.335*(1 -0.335))))

#Finding P(21 <= X <= 24) with the above mean and SD
(pnorm(24,  mean = HTnormVals[1], sd = HTnormVals[2]) - pnorm(20, mean = HTnormVals[1], 
                                                              sd = HTnormVals[2])) %>%
  round(4)
```

The probability in this case, 0.3996 (rounded to 0.4), is very similar to the binomial probability in Question 6.3, indicating that the population in question is most likely normally distributed. 
