---
title: "My answers"
author: "My name"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

# Question 1: Blood types probability
The following table shows the prevalence of ABO block groups for the U.S. Caucasian and American Black populations. The table specifies the probabilities of a specific race and blood type. For example, the probability that a randomly selected individual is black with blood type O is 0.098.


|                |    A     |    B     |     AB  |      O    |
| -------------- |:--------:|:--------:|:--------:|:--------:|
| Caucasian      |   0.352  |   0.064  |   0.024  |  0.360   |
| American Black |   0.054  |   0.040  |   0.008  |  0.098   |

## Question 1.1
What is the probability that a randomly selected individual does not have blood type AB?

```{r}
# Write your answer here
```


## Question 1.2
What is the probability that two people selected at random both have blood type B?

```{r}
# Write your answer here
```


## Question 1.3
Are the events "blood type B" and "American black" race statistically independent?

```{r}
# Write your answer here
```


## Question 1.4
Are the events "blood type O" and "blood type A" mutually exclusive?

```{r}
# Write your answer here
```


## Question 1.5
Given an example of two events that are not mutually exclusive.

```{r}
# Write your answer here
```


## Question 1.6
What is the conditional probability of "blood type A" given that the race is "American black"?

```{r}
# Write your answer here
```


# Question 2: Death probability
The following data records the reported proportions of 20- to 25-year-old White males causes of death within a five year period. In other words, it gives the probability of death within five years.


|  Cause           |    Probability  |
| ---------------- |-----------------:|
| Suicide          |   0.00126        |   
| Homicide         |   0.00063        | 
| Auto accident    |   0.00581        |
| Leukemia         |   0.00023        |
| all other causes |   0.00788        |

## Question 2.1
What is the probability that a White male aged 20 to 25 dying from any cause of death within the next five years?

```{r}
# Write your answer here
```


## Question 2.2
Out of 100,000 white males in the 20 to 25 age group, how many deaths would you expect in the next 5 years from leukemia?

```{r}
# Write your answer here
```


## Question 2.3
Given that a White male aged 20 to 25 years has died, what is the most likely cause of death? Assume that nothing else is known. Provide rationale for your answer.

```{r}
# Write your answer here
```


# Question 3: Lab test one
Suppose that a lab test values are normally distributed with $\mu = 80$ and $\sigma = 3$. 

## Question 3.1
What is the probability that a randomly selected test has a value that is within one standard deviation of the mean? Round to three decimal places.

```{r}
# Write your answer here
```


```{r}
# Write your answer here
```


## Question 3.2
What is the probability that a randomly selected test has a value that is within two standard deviations of the mean? Round to three decimal places.

```{r}
# Write your answer here
```


```{r}
# Write your answer here
```


## Question 3.3
What is the probability that a randomly selected test has a value that is within three standard deviations of the mean? Round to three decimal places.

```{r}
# Write your answer here
```


```{r}
# Write your answer here
```


# Question 4: Lab test two
Suppose that another lab test values record changes from a reference value and these are normally distributed with $\mu = -16$ and $\sigma = 1.5$. 

## Question 4.1
What is the probability that a randomly selected test has a value that is within one standard deviation of the mean? Round to three decimal places.

```{r}
# Write your answer here
```


```{r}
# Write your answer here
```


## Question 4.2
What is the probability that a randomly selected test has a value that is within two standard deviations of the mean? Round to three decimal places.

```{r}
# Write your answer here
```


```{r}
# Write your answer here
```


## Question 4.3
What is the probability that a randomly selected test has a value that is within three standard deviations of the mean? Round to three decimal places.

```{r}
# Write your answer here
```


```{r}
# Write your answer here
```


## Question 4.4
How do your answers compare to those of Question 3? Does this surprise you?

```{r}
# Write your answer here
```


# Question 5: Randomization verification
A clinical trial was designed to determine whether radiation therapy or surgery was more effective in reducing disease recurrence in men with prostate cancer. Patients were randomized to each treatment in a 1:1 manner, which implies the chance a patient is randomized to either arm is 0.50. Suppose that in the first 20 patients randomized at a particular site participating in the trial, 17 patients were assigned to the surgery arm. The site complains that the randomization does not appear to be working. Does the site's suspicion have merit? 

```{block,  box.title = "Tip", box.body = list(fill = "lemonchiffon")}
When you answer this, you should consider all scenarios that would support the suspicions that the randomization is broken. Specifically, what other outcomes out of 20 would be more extreme to the one observed. You might want to calculate the probability of getting the result observed or one more extreme.
```

```{r}
# Write your answer here
```


```{r}
# Write your answer here
```


```{r}
# Write your answer here
```


# Hypertension
The remainder of the questions are relevant to this scenario. 

The proportion of U.S. adults over the age of 20 with hypertension is 0.335, or 33.5%. 

# Question 6: Hypertension
Suppose you take a random sample of 64 U.S. adults over the age of 20.

## Question 6.1
What is the chance that 32 people have hypertension? (Round to 4 decimal places.)

```{r}
# Write your answer here
```


```{r}
# Write your answer here
```


## Question 6.2
What is the chance that more than 20 individuals in the sample have hypertension? (Round to 3 decimal places.)

```{r}
# Write your answer here
```


```{r}
# Write your answer here
```


## Question 6.3
What is the chance that between 21 and 24 individuals, inclusive, out of 64 have hypertension? (Round to 3 decimal places.)

```{r}
# Write your answer here
```


```{r}
# Write your answer here
```


# Question 7: Hypertension
Now suppose you take your random sample (of size 64) from one city in the United States. You get a sample proportion of 0.211.


## Question 7.1
How likely are you to get a sample proportion this small or smaller if this city is representative of the US population?

```{r}
# Write your answer here
```


```{r}
# Write your answer here
```


## Question 7.2
What do you conclude?

```{r}
# Write your answer here
```


# Question 8: Hypertension
Give a range for which you are 95% certain would contain the sample proportion of adults with hypertension if you were to take a random sample of 64 adults in the U.S. over the age of 20. (Round to three decimal places.)

```{r}
# Write your answer here
```


```{r}
# Write your answer here
```


```{r}
# Write your answer here
```


# Question 9: Hypertension
We want to compare the sampling distribution of our sample proportion to that of a normal distribution. The normal distribution should have the same mean and standard deviation as the sampling distribution. To do this, we want to use `qqplot()`. We will simulate the sampling distribution of both the sampling proportion as well as the normal distribution. After using `qqplot()`, use `qqline()` to add a line to the plot.

## Question 9.1
Make a qqplot comparing the sampling distribution of the sample proportion to that of a normal distribution, including a line that indicates a perfect relationship. 

```{r}
# Write your answer here
```


```{r}
# Write your answer here
```


## Question 9.2
How do these two distributions compare?

```{r}
# Write your answer here
```


## Question 9.3
To verify your answer to Question 9.2, use a normal distribution with the same mean and standard deviation of the binomial distribution to determine the likelihood that we observe between 21 and 24 individuals, inclusive, out of 64 have hypertension (Question 6.3). How do these values compare?

```{r}
# Write your answer here
```


```{r}
# Write your answer here
```


```{r}
# Write your answer here
```

